\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-04-21 mSystems/mSys_MandM.tex
mSystems release 0.0.4
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_results.tex
	modified:   results/finalPlot33_new.pdf

We summarize our dataset in ST1-6 (Tables 1-6 in suplemental material). The bacteria and archaea taxonomic assignations were obtained by analysing 16S rRNA sequences, which were clustered into operational taxonomic units (OTUs) sharing 97 \% sequence identity using QIIME\cite{QIIME}. WGS data\cite{kwashiorkor} were analysed and assigned at strain level by the Livermore Metagenomic Analysis Toolkit (LMAT)\cite{LMAT}, according to their default quality threshold. Genus, with best balance between error assignment and number of taxa, was chosen as our reference taxonomic level. We have verified that our conclusions are not significantly affected by selecting family or species as the reference taxonomic level (see Figure 1 in supplemental material).

\todo{Specify, in each study treated, the nature of the samples (conditions, timespan between timepoints, subjects). Specify, and it is very important, what we consider ?healthy? in each study (for example: pre-antibiotics is healthy)}

@@

@@

The effective data analysis has been performed with a Python tool developed from scratch to more than 4200 lines of code. Implemented following the Object Oriented Programming, paradigm, this software is the back-end of the website described above. However, it could be run as an independent piece of software since it is built as a Python package provided with a command-line front-end (\CC.py). Once installed, the tool can be run interactively but also in automatic mode, which uses parallel computation to speed up the analysis of several data sources.

@@

The basic idea of bootstrapping is that inference about a population from sample data (sample $\rightarrow$ population) can be modeled by resampling the sample data and performing inference on (resample $\rightarrow$ sample). To adapt this general idea to our problem, we resample the x-data array using its errors array. That is, for each replicate, a new x-data array is computed based on:
\begin{linenomath}
$$x^*_i = x_i + v_i$$
\end{linenomath}
where $v_i$ is a Gaussian random variable with mean $\mu_i=0$ and standard deviation $\sigma_i=\mathrm{SEM}_i$, as defined previously in this supplementary material. For each replicate a complete un-weighted power-law fit is performed, as described in the previous section. It is worth mentioning that each replicate is filtered to avoid values of $x^*_i$ under \emph{eps} (obtained by \texttt{np.finfo(np.double).eps}) in order to keep away from the error of getting log of negatives or zero during the fit.

>>>>

@@

The bacteria and archaea taxonomic assignations were obtained by analysing 16S rRNA sequences, which were clustered into operational taxonomic units (OTUs) sharing 97 \% sequence identity using QIIME\cite{QIIME}. WGS data\cite{kwashiorkor} were analysed and assigned at strain level by the Livermore Metagenomic Analysis Toolkit (LMAT)\cite{LMAT}, according to their default quality threshold. Genus, with best balance between error assignment and number of taxa, was chosen as our reference taxonomic level. We have verified that our conclusions are not significantly affected by selecting family or species as the reference taxonomic level (see Figure \ref{fig:taxlev1}).

\todo{Specify, in each study treated, the nature of the samples (conditions, timespan between timepoints, subjects). Specify, and it is very important, what we consider \emph{healthy} in each study (for example: pre-antibiotics is healthy)}

@@

The effective data analysis has been performed with a Python tool developed from scratch to more than 4200 lines of code. Implemented following the Object Oriented Programming paradigm, this software is the back-end of the website described above. However, it could be run as an independent piece of software since it is built as a Python package provided with a command-line front-end (\CC.py). Once installed, the tool can be run interactively but also in automatic mode, which uses parallel computation to speed up the analysis of several data sources.

@@

The basic idea of bootstrapping is that inference about a population from sample data (sample $\rightarrow$ population) can be modeled by resampling the sample data and performing inference on (resample $\rightarrow$ sample). To adapt this general idea to our problem, we resample the x-data array using its errors array. That is, for each replicate, a new x-data array is computed based on:
\begin{linenomath}
$$x^*_i = x_i + v_i$$
\end{linenomath}
where $v_i$ is a Gaussian random variable with mean $\mu_i=0$ and standard deviation $\sigma_i=\mathrm{SEM}_i$, as defined previously. For each replicate a complete un-weighted power-law fit is performed, as described in the previous section. It is worth mentioning that each replicate is filtered to avoid values of $x^*_i$ under \emph{eps} (obtained by \texttt{np.finfo(np.double).eps}) in order to keep away from the error of getting log of negatives or zero during the fit.

### 2016-04-21 mSystems/mSys_results.tex
mSystems release 0.0.4
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_results.tex
	modified:   results/finalPlot33_new.pdf

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Tables \label{tab:diet} to \label{tab:LEA}. Our results hint at an ubiquotous behaviour. Firstly, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota, and the scaling index is always smaller than one, which means that, more abundant taxa are less volatile than less abundant ones. Secondly, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define the healthy zone in the Taylor parameter space.

@@

Taylor's power law has been explained in terms of various effects, all without general consensus. It can be shown to have its origin in a mathematical convergence similar to the central limit theorem, so virtually any statistical model designed to produce a Taylor law converge to a Tweedie distribution\cite{stat}, providing a mechanistic explanation based on the statistical theory of errors\cite{convergence1,convergence2,convergence3}. To unveil the generic mechanisms that drive different scenarios in the $\beta$--$V$ space, we model the system by assuming that taxon relative abundance follows a Langevin equation with a deterministic term that captures the fitness of each taxon and a randomness term with Gaussian random noise\cite{ranking}. Both terms are modeled by power laws, with coefficients that can be interpreted as the taxon fitness $F_i$ and the variability $V$ (see Model in Material and Methods). When $V$ is sufficiently low, abundances are stable in time.  Differences in variability $V$ can induce a noise-induced phase transition in relative abundances of taxa. The temporal evolution of the probability of a taxon having abundance $x_i$ given its fitness is governed by the Fokker--Planck equation. The results of solving this equation show that the stability is best captured by fitness $F$ and amplitude of fluctuations $V$ phase space (see Figure \ref{fig:main3}).

\begin{figure}
	\centering
	\includegraphics[width=1.0\textwidth]{results/comboplot2.pdf}
	\caption{Phase space of microbiota stability. We show the Variability parameter as a function of time for the male in Caporaso's study\cite{moving} (Figure A) and for the patient D in the antibiotics study\cite{antibiotic} (Figure B). Periods of antibiotic treatment are shown by blue vertical lines. Microbiota states can be placed in the phase space $F$--$V$ (Figure C). The light blue shaded region corresponds to the stable phase, while the grey shaded region is the unstable phase (the phase transition line is calculated for  $\alpha$ = $\beta$ = 0.75). We place healthy individuals (green) and individuals whose gut microbiota is threatened (antibiotics, IBS) in the phase space fitness--variability. Gut microbiota of healthy individuals over a long term span show a quasi--periodical variability (central period is ten days). We show that taking antibiotics (AB1 and AB2 correspond to first and second treatment respectively) induces a phase transition in the gut microbiota, which impacts its future changes. We also show an IBS--diagnosed patient transiting from the unstable to the stable phase.}
	\label{fig:main3}
\end{figure}

@@

>>>>

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Tables \ref{tab:diet} to \ref{tab:LEA}. Our results hint at an ubiquotous behaviour. Firstly, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota, and the scaling index is always smaller than one, which means that, more abundant taxa are less volatile than less abundant ones. Secondly, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define the healthy zone in the Taylor parameter space.

@@

Taylor's power law has been explained in terms of various effects, all without general consensus. It can be shown to have its origin in a mathematical convergence similar to the central limit theorem, so virtually any statistical model designed to produce a Taylor law converge to a Tweedie distribution\cite{stat}, providing a mechanistic explanation based on the statistical theory of errors\cite{convergence1,convergence2,convergence3}. To unveil the generic mechanisms that drive different scenarios in the $\beta$--$V$ space, we model the system by assuming that taxon relative abundance follows a Langevin equation with a deterministic term that captures the fitness of each taxon and a randomness term with Gaussian random noise\cite{ranking}. Both terms are modeled by power laws, with coefficients that can be interpreted as the taxon fitness $F_i$ and the variability $V$ (see Model under Material and Methods). When $V$ is sufficiently low, abundances are stable in time.  Differences in variability $V$ can induce a noise-induced phase transition in relative abundances of taxa. The temporal evolution of the probability of a taxon having abundance $x_i$ given its fitness is governed by the Fokker--Planck equation. The results of solving this equation show that the stability is best captured by fitness $F$ and amplitude of fluctuations $V$ phase space (see Figure \ref{fig:main3}).

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/finalPlot33_new.pdf}
	\caption{Microbiota states can be placed in the phase space $F$--$V$. The light blue shaded region corresponds to the stable phase, while the grey shaded region is the unstable phase (the phase transition line is calculated for  $\alpha$ = $\beta$ = 0.75). We place healthy individuals (green) and individuals whose gut microbiota is threatened (antibiotics, IBS) in the phase space fitness--variability. Gut microbiota of healthy individuals over a long term span show a quasi--periodical variability (central period is ten days). We show that taking antibiotics (AB1 and AB2 correspond to first and second treatment respectively) induces a phase transition in the gut microbiota, which impacts its future changes. We also show an IBS--diagnosed patient transiting from the unstable to the stable phase.}
	\label{fig:main3}
\end{figure}

@@

Finally, we have studied the time dependence of the variability $V$ and power law index $\beta$ (see Model under Material and Methods) by using a sliding window approach. The total number of time points are divided in subsets of five points, where next subset is defined by adding next time sampling and by eliminating the earliest one. Both parameters were calculated for each subset against the average time lapse. Figure \ref{fig:tempevo1} shows the variability  $V$ as a function of time for the largest sampling: two individuals in the Caporaso's study\cite{moving} corresponding to the gut microbiota of a male (upper plot) and a female (lower plot). Figure \ref{fig:tempevo2} shows the time evolution of $V$ for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot).

\begin{figure}
	\includegraphics[width=1.0\textwidth]{results/sliwin/male_mov.pdf}
	\hspace*{3mm}\includegraphics[width=0.448\textwidth]{results/sliwin/female_mov.pdf}
\caption{$V$ as a function of time for the two individuals in the Caporaso's study\cite{moving}: samples of gut microbiome of a male (upper plot) and a female (lower plot). Both samples show changes in the variability V with quasi--periodic behavior peaked at about 10 days. Variability grows more for the gut microbiota of the male and share a minimal value around 0.1 with the gut microbiota of the female.}
\label{fig:tempevo1}
\end{figure}

\begin{figure}
	\centering 
 	\includegraphics[width=0.8\textwidth]{results/sliwin/patP2_IBS.pdf}
  	\includegraphics[width=1.0\textwidth]{results/sliwin/patD_antibio.pdf} 
\caption{$V$ as a function of time for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot). The variability 
of the gut microbiota of P2 decreases from above 0.3 to below 0.2, showing a slow tendency to increase the order of the system.  Antibiotic intake leaks to a quick increase of variability which lasts for a few days to recover ordering. The second antibiotic treatment shows some memory (lower increase of variability) with a slower recovery. NOTE: The blue vertical lines in the lower plot are showing the periods of antibiotic treatment.}
\label{fig:tempevo2}
\end{figure}
